---
title: "Find labeled"
author: "Chuyao Wang"
date: "`r Sys.Date()`"
output: 
  html_document: 
    number_sections: true # automatic section numbering
    toc: yes # table of content
    toc_depth: 3
    toc_float: true
    smooth_scroll: true
    toc_collapsed: true
    fig_caption: yes
    df_print: kable
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "./html") })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(xcms)
# library(CAMERA)
library(dplyr)
library(stringr)
library(RColorBrewer)
library(pander)
library(openxlsx)
library(ggplot2)
library(hrbrthemes)
library(gtools)
library(readr)
library(GDAtools)
source("isotopeApp/helpers.R")
```

```{r serial-setup}
register(SerialParam())
```

# Data import

```{r loadData}
datadir <- "./data/20230228"
file <- list.files(path = datadir, pattern = "^.*\\.mzML$",
                   all.files = TRUE, full.names = TRUE,
                   recursive = FALSE, ignore.case = FALSE,
                   include.dirs = FALSE, no.. = TRUE) %>%
  mixedsort(decreasing=T)
pd <- data.frame(file = basename(file),
                 sample = c("Sample1","Sample2"),
                 group = c("SQ","JN"))
data <- readMSData(file, # filenames to load
                   # Stores data; stores sample info into @phenoData@data
                   # More info about each variable: file, injection. sample etc. can be entered in varMetadata
                   # pdata = new("NAnnotatedDataFrame", pd),
                   mode = "onDisk" # Only load metadata into memory and leave the m/z ratio vs intensity peaks on disk
) %>% filterMsLevel(1)
```

```{r viewData}
# featureData <- fData(data)
# 
# polarity(data) %>% table
# msLevel(data) %>% table
```

# Data centroiding

```{r denoise-and-centroid, results='hide', warning=FALSE}
data_cent <- data %>%
      smooth(method = "SavitzkyGolay", halfWindowSize = 10) %>%
      #combineSpectraMovingWindow() %>%
      pickPeaks(refineMz = "descendPeak") 

name <- fileNames(data) %>%
  str_remove(".mzML") %>%
  paste("cent",sep = "_") %>%
  paste("mzML",sep = ".")

# lapply(name, function (z) {
#     if (file.exists(z))
#         file.remove(z)
# })

writeMSData(data_cent,file=name)

```

# Pre-processing

## Read the centroided data

```{r}
datadir <- "./data/20230228"

file <- list.files(path = datadir, pattern = "cent",
                   all.files = TRUE, full.names = TRUE,
                   recursive = FALSE, ignore.case = FALSE,
                   include.dirs = FALSE, no.. = TRUE) %>%
  mixedsort(decreasing = TRUE) %>% 
  `[`(1:2)

pd <- data.frame(file = basename(file),
                 sample = c("Sample1","Sample2"),
                 group = c("SQ","JN"))

data_cent <- readMSData(file, pdata = new("NAnnotatedDataFrame", pd),
                        mode = "onDisk")  
```

## Peak detection

centwavewithprediso: do one centwave run, predict potential isotope positions, do another centwave in those positions

centwave: identify roi first, then do cwt in those rois to find peaks.

roi: region in the mz-rt space where consecutive mz values deviate less than ppm

### Fine-tune parameters

#### Peak width

```{r, warning=F}
data %>%
  filterFile(2) %>%
  filterMz(mz=get_ppm_range(90.05496,10)) %>%
  filterRt(rt=c(390,420)) %>%
  plot(type="XIC")
```

> about 50s

```{r, warning=F}
data_cent %>%
  filterFile(2) %>%
  filterMz(mz=get_ppm_range(90.05496,5)) %>%
  filterRt(rt=c(390,420)) %>%
  chromatogram(missing=0) %>%
  plot
```

1 min, 40s

### Set parameters

```{r}
# params <- CentWavePredIsoParam(ppm = 10,
#                                peakwidth = c(20,70),
#                                snthresh = 30,
#                                prefilter = c(3,7e5),
#                                mzCenterFun = "wMeanApex3",
#                                integrate = 2,
#                                mzdiff = -0.001, # if negative allow overlap, if positive and <0.001, count as one peak, if >0.001, count as two peaks
#                                noise = 3e5,
#                                maxCharge = 1,
#                                maxIso = 5,
#                                polarity = "negative"
# )

params <- CentWaveParam(ppm = 5,
                        peakwidth = c(3,310),
                        snthresh = 10,
                        prefilter = c(3,1),
                        mzCenterFun = "wMeanApex3",
                        integrate = 1,
                        mzdiff = -0.001, # if negative allow overlap, if positive and <0.001, count as one peak, if >0.001, count as two peaks
                        noise = 10000

)
group_colors <- brewer.pal(length(fileNames(data_cent)),"Set2")
```

### Test parameters

```{r, warning=FALSE}
chr_glu <- data_cent %>%
  filterFile(2) %>%
  filterMz(mz=get_ppm_range(90.05496,5)) %>%
  filterRt(rt=c(390,420)) %>%
  chromatogram(missing=0)
plot(chr_glu, col=group_colors)
```

Test on XIC

```{r}
params.test <- params
params.test@snthresh <- 2
xchr <- findChromPeaks(chr_glu,params.test,BPPARAM=SerialParam())
```

```{r}
plot(xchr,col=group_colors,peakBg=group_colors[chromPeaks(xchr)[, "column"]])
```

### Do peak picking

```{r, warning=FALSE}
xdata <- findChromPeaks(data_cent, param = params, return.type = "XCMSnExp",BPPARAM=SnowParam())
xdata_og <- xdata
```

## Peak refinement

### Peak merging

> Reduced many peaks for predIsoparams method

```{r warning=FALSE}
mpp <- MergeNeighboringPeaksParam(expandRt = 2, expandMz=0, ppm = 5, minProp=0.75)
xdata_mpp <- refineChromPeaks(xdata, param = mpp, BPPARAM=SerialParam())
```

### Get peak summaries

```{r}
xdata <- xdata_mpp
```

#### Summary table

```{r}
summary_fun <- function(z)
    c(peak_count = nrow(z), rt = quantile(z[, "rtmax"] - z[, "rtmin"]))

T <- lapply(split.data.frame(
    chromPeaks(xdata), f = chromPeaks(xdata)[, "sample"]),
    FUN = summary_fun)
T <- do.call(rbind, T)
rownames(T) <- basename(fileNames(xdata))
pandoc.table(
    T,
    caption = paste0("Summary statistics on identified chromatographic",
                     " peaks. Shown are number of identified peaks per",
                     " sample and widths/duration of chromatographic ",
                     "peaks."))
```

#### Location of peaks per file

```{r, warning=F}
plotChromPeaks(xdata, file =2)
```

#### Number of peaks along rt axis

```{r}
plotChromPeakImage(xdata,yaxt = "n")
```

#### Peak intensities per file

```{r}
ints <- split(log2(chromPeaks(xdata)[, "maxo"]),
              f = chromPeaks(xdata)[, "sample"])
boxplot(ints, varwidth = TRUE, col = group_colors[xdata$sample_group],
        ylab = expression(log[2]~intensity), main = "Peak intensities")
grid(nx = NA, ny = NULL)
```

## Retention time correction

```{r}
register(SerialParam())
xdata <- xdata_mpp
xdata_rt <- adjustRtime(xdata, param = ObiwarpParam(
  gapInit = 0,
  gapExtend = 2.7,
  factorDiag = 2,
  factorGap = 1,
  localAlignment = FALSE,
  binSize = 0.18, # profStep in retcor
  response=19.25,
  distFun = "cor_opt"
  ))
```

```{r}
par(mfrow = c(2,1))

chr_glu_before <- chromatogram(xdata_mpp,mz=get_ppm_range(185.0762,5),rt=c(380,420))
plot(chr_glu_before, col=group_colors, peakType="none")
chr_glu_after <- chromatogram(xdata_rt,mz=get_ppm_range(185.0762,5),rt=c(380,420))
plot(chr_glu_after,col=group_colors,peakType="none")
```

## Peak grouping

```{r}
xdata <- xdata_rt
```

### Test parameters

Group similar peaks in each file into a feature

```{r}
## Define the mz slice.
mzr <- get_ppm_range(185.0762,5)

## Extract and plot the chromatograms
chr_mzr <- chromatogram(xdata, mz = mzr,rt=c(380,440))
## Define the parameters for the peak density method
pdp <- PeakDensityParam(sampleGroups = c(1,1,1,1),
                        bw = 5, # adjust according to result
                        minFraction = 0.5,
                        minSamples = 2, 
                        binSize = 0.00415, #mzwid
                        maxFeatures = 50
                        )
plotChromPeakDensity(chr_mzr, col = group_colors, param = pdp,
                     peakBg = group_colors,
                     peakCol = group_colors,
                     peakPch = 16)
```

All glucose peaks are grouped into one feature

### Do peak grouping

```{r}
register(SerialParam())
xdata_gpk <- groupChromPeaks(xdata, param = pdp)
```

## Peak filling

Fill in missing peaks from each feature

```{r}
xdata <- xdata_gpk
```

```{r}
register(SerialParam())
xdata_fill <- fillChromPeaks(xdata, param = ChromPeakAreaParam())
```

# Targeted search

## Extract peaklist

```{r}
xdata <- xdata_mpp
peaklist <- chromPeaks(xdata) %>% as.data.frame
```

## Read metabolites table

```{r}
# compounds <- read.xlsx("data/metabolites_neg.xlsx",sheet=1,startRow = 1)
compounds <- read.xlsx("data/metabolites_pos.xlsx",sheet=1,startRow = 1)
elements <- c("C","H","N","O","P","S")
count <- matrix(nrow=length(compounds$Formula),ncol = length(elements)) %>% as.data.frame
for (i in seq_along(elements)) {
  element_notexist <- str_extract(compounds$Formula,elements[i]) %>% is.na
  element_number <- str_extract(compounds$Formula,paste(elements[i],"\\d+",sep="")) %>%
    str_extract(.,"\\d+") %>%
    as.numeric()
  element_number[element_notexist] <- 0
  element_number[is.na(element_number)] <- 1
  
  count[,i] <- element_number
}
colnames(count) <- elements

count <- mutate(count, ion_formula = paste("C",C,"H",H+1,"N",N,"O",O,"P",P,"S",S,sep="")) # change to H+1 for positive mode
for (i in seq_along(count$ion_formula)) {
  vec <- str_extract_all(count$ion_formula[i],"\\D?\\d+") %>% unlist 
  vec[str_detect(vec,"^\\D?0$")] <- ""
  vec[str_detect(vec,"^\\D?1$")] <- str_remove(vec[str_detect(vec,"^\\D?1$")],"1")
  count$ion_formula[i] <- str_flatten(vec)
}
compounds <- bind_cols(compounds, count)
```

```{r}
peaklist <- read.csv("data/20230721/peaklist.csv")
```

## Filter for mz values

```{r}
abbs <- compounds$Abbreviation
isocor <- data.frame()

mdv <- matrix(nrow = length(compounds$Name),ncol=length(unique(peaklist$sample))) %>%
  as.data.frame

for (spl in unique(peaklist$sample)) {
  for (cpd_idx in seq_along(compounds$Name)) {
    num_labeled <- 6 # change to max # of labeled carbon atoms
    row <- compounds[cpd_idx,]
    if (row$C<num_labeled) {
      num_labeled <- row$C
    }
    mzs <- get_mz_ch(C=row$C,
                     N=row$N,
                     H=row$H,
                     S=row$S,
                     O=row$O,
                     P=row$P,
                     c13=num_labeled, 
                     d=0,
                     pol=1 ) %>% # -1 for neg mode
      unlist %>% 
      rev
    
    extracted <- data.frame(area=c(),rt=c(),isotopologue=c()) # stores area for each mz value
    for (mz_idx in seq_along(mzs)) { # going from M+0 to M+num_labeled
      temp <- peaklist %>%
        filter(sample == spl) %>% 
        filter(abs(mz-mzs[mz_idx])<0.001) %>% 
        select(into,rt) %>% 
        mutate(label = mz_idx-1)
      if (nrow(temp)==0) { # If no peak is found, add a dummy peak
        temp[nrow(temp)+1,] <- c(0,0,mz_idx-1) # Add dummy peak
      }
      colnames(temp) <- c("area","rt","isotopologue")
      extracted <- bind_rows(extracted,temp)
    }
    
    if (sum(extracted$area)==0) { # If not found at all
      mdv[cpd_idx,spl] <- rep(0,times=num_labeled) %>% as.character() %>% str_flatten(.,collapse = " ") # MDV is all 0s
    } else {
    # Remove peaks with inconsistent rt from the rest of the peaks and low areas
    extracted <- filter_deviation(extracted,rt_thresh = 5,area_thresh = 1e4)
    areas <- extracted$area
    
    mdv[cpd_idx,spl] <- (areas/sum(areas)) %>% round(.,digits=4) %>% as.character() %>% str_flatten(.,collapse = " ") # MDV is calculated
      
      isocor.sample <- rep(paste("Sample_",spl,sep=""),times=length(areas))
      isocor.metabolite <- rep(abbs[cpd_idx],times=length(areas))
      isocor.derivative <- rep(NA,times=length(areas))
      isocor.isotopologue <- 0:(length(areas)-1)
      isocor.area <- areas
      isocor.resolution <- rep(120000,length(areas))
      
      isocor.add <- data.frame(isocor.sample,isocor.metabolite,isocor.derivative,isocor.isotopologue,isocor.area,isocor.resolution)
      
      isocor <- bind_rows(isocor,isocor.add)
    }
  }
}
# colnames(mdv) <-  fileNames(xdata) %>% basename
compounds <- bind_cols(compounds,mdv)

colnames(isocor) <- c("sample","metabolite","derivative","isotopologue","area","resolution")
```

## Make metabolite.dat

```{r}
metdat <- compounds %>% 
  select(Abbreviation, ion_formula) %>% 
  mutate(charge=rep(1,length(Abbreviation))) %>% 
  mutate(inchi=rep(NA,length(Abbreviation)))
colnames(metdat)[1:2] <- c("name","formula")
```

## Save data

```{r}
# For viewing results
write_csv(compounds,"output/mdv0228_pos2.csv",na="")
# For isocor
write_tsv(isocor,"output/isocordata0228_pos2.tsv",na="")
# For isocor database
write_tsv(metdat,"output/Metabolites.dat",na="",quote = "none")
```

## Process result

```{r}
# For INCA
# Read isocor corrected result
res <- read_tsv("D:/Files/SH/isocor/0228_pos2/isocordata0228_pos2_res.tsv")
res <- res %>%
  filter(sample=="Sample_ 2") %>% 
  select(metabolite,isotopologue_fraction) %>% 
  mutate(error = rep(0.02,times=nrow(.))) %>% 
  split.data.frame(.,.$metabolite)

add <- data.frame(metabolite="",isotopologue_fraction=as.integer(0),error=as.numeric(""))
res_new <- data.frame()
for (i in 1:length(res)) {
  res_new <- bind_rows(res_new,add,res[[i]])
}

# Save file for import into INCA
write_csv(res_new,"D:/Files/SH/isocor/0228_pos2/res_processed_pos2.csv",na="")
```

# Data format conversion

```{r}
# data_set <- as(xdata_fill,"xcmsSet")
```

> sampclass is a method to add or adjust the class column in the phenoData slot of a xcmsSet object

# Isotope analysis

intb: baseline corrected area into: area maxo: peak intensity

```{r}
an <- xsAnnotate(data_set, polarity="negative") # constructor; extracts peak table
an <- groupFWHM(an, perfwhm = 10,intval = "maxo") # group peaks by retention time
an <- findIsotopes(an,
                    ppm=15,
                    mzabs=0.1,
                    intval="maxo",
                    maxcharge = 1,
                    maxiso=8,
                    filter=TRUE)
an <- groupCorr(an,cor_eic_th=0.5,pval=0.1)
an <- findIsotopes(an,
                    ppm=5,
                    mzabs=0.001,
                    intval="maxo",
                    maxcharge = 1,
                    maxiso=8,
                    filter=TRUE)
an <- findAdducts(an,ppm=5,mzabs=0.001,multiplier=2,polarity = "negative",intval="maxo")
```

## Get result

```{r}
peaklist <- getPeaklist(an) # extract peak list
```

do peak merging annotation
