---
title: "metDNA processing"
author: "Chuyao Wang"
date: "`r Sys.Date()`"
output: 
  html_document: 
    number_sections: true # automatic section numbering
    toc: yes # table of content
    toc_depth: 3
    toc_float: true
    smooth_scroll: true
    toc_collapsed: true
    fig_caption: yes
    df_print: kable
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "./html") })
---

# Description

Prepare the input table for metaboanalyst

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(stringr)
library(gtools)
library(dplyr)
library(stringdist)
strfil_colnames <- function(df,pattern) {
  a <- colnames(df)[str_detect(colnames(df),pattern)]
  return(a)
}
dir.anno.pos <- "~/Documents/metDNA/metDNA Output/cellsamples_4genes_pos/POS/00_annotation_table/table3_identification_pair.csv"
dir.anno.neg <- "~/Documents/metDNA/metDNA Output/cellsamples_4genes_neg/00_annotation_table/table3_identification_pair.csv"
dir.peak.pos <- "~/Documents/metDNA/data/pos/PeakTable.csv"
dir.peak.neg <- "~/Documents/metDNA/data/neg/PeakTable.csv"
```

# Load data
anno: annotated peaks from metDNA2

peak: original peak table from xcms; missing values not imputed; contain areas for qc samples

Since peak names can be similar across positive and negative peak tables. An "n" is appended before the peak names in negative mode peaks

```{r}
anno.pos <- read.csv(dir.anno.pos) %>%
  mutate(peak_name2 = str_remove(peak_name,"_[:letter:]")) %>%
  select(peak_name,peak_name2,everything())
peak.pos <- read.csv(dir.peak.pos)
colnames(anno.pos) <- colnames(anno.pos) %>% str_remove(.,"_Pos")
colnames(peak.pos) <- colnames(peak.pos) %>% str_remove(.,"_Pos")

anno.neg <- read.csv(dir.anno.neg) %>%
  mutate(peak_name = str_c("n",peak_name)) %>%
  mutate(peak_name2 = str_remove(peak_name,"_[:letter:]")) %>%
  select(peak_name,peak_name2,everything())
peak.neg <- read.csv(dir.peak.neg) %>%
  mutate(name = str_c("n",name)) %>%
  select(name,everything())
colnames(anno.neg) <- colnames(anno.neg) %>% str_remove(.,"_Neg")
colnames(peak.neg) <- colnames(peak.neg) %>% str_remove(.,"_Neg")
```

# Impute missing values with RF
Impute the missing values in the peak table with random forest

```{r}

```


# Perform SERRF QC correction
Correct the peak areas using QC areas as reference

# Remove duplicate annotations in the anno table
The annotations should be filtered in the postive and negative modes combined.
```{r}
anno <- bind_rows(anno.pos,anno.neg)
```

First metabolites with duplicate names should be removed according to the name column in the anno table.
```{r}
not_max <- function(x) {
  not.max <- which(x!=max(x))
  return(not.max)
}

anno <- anno %>%
  mutate(total.area = anno %>%
           select(colnames(anno)[str_detect(colnames(anno),"FS")]) %>%
           apply(.,MARGIN=1,sum))

duped <- anno$name %>% table %>% as.data.frame() %>% `[`(.$Freq>1,) %>% `[`(,1) %>% as.vector()
```

Besides identical metabolites, stereoisomers, such as D-Leucine and L-Leucine should also be removed. This requires computing a distance matrix of the smiles names of the metabolites since smiles name also indicate structure. Structurally similar compounds will have a lower distance. The matrix can then be filtered given a cutoff.

Many compounds have smiles name but not common name. So unique(smiles)>unique(name). It is also possible for a smile name to be related to multiple common names, since smile names refer to structures. There is also the case where a compound has multiple smiles name, such as creatine.

```{r}
anno <- bind_rows(anno.pos,anno.neg)
# distmat <- stringdistmatrix(unique(anno$smiles), unique(anno$smiles), "jw", p=0.1)
distmat <- readRDS("./outputs/distmat4metdna.rds")
```


```{r}

```

Use the distance matrix to filter



# Filter out peaks with the same annotations
Multiple peaks can have the same annotations because of different adducts or retention times. These are dependent variables and should be removed to only keep the most relevant one.


# Data cleaning

## Low intensity values (mean/median)

## Constant values (sd/iqr)

## High RSD in QC samples

# Remove duplicate metabolites

# Normalization
