---
title: "xcms tutorial"
author: "Chuyao Wang"
date: "`r Sys.Date()`"
output: 
  html_document: 
    number_sections: true # automatic section numbering
    toc: yes # table of content
    toc_depth: 3
    toc_float: true
    smooth_scroll: true
    toc_collapsed: true
    fig_caption: yes
    df_print: kable
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "./html") })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(xcms)
library(MSnbase)
library(msdata)
library(magrittr)
library(png)
library(dplyr)
library(knitr)
```

# Mass spectrometry

Mass spectrometry allows to measure abundances of charged molecules (ions) in a sample. Abundances are determined as ion counts for a specific mass-to-charge ratio m/z. The measured signal is represented as a spectrum: intensities along m/z.

In such GC/LC-MS based untargeted metabolomics experiments the data is analyzed along the retention time dimension and chromatographic peaks (which are supposed to represent the signal from a ion species) are identified and quantified.

MSnbase: used for data handling (data import, access, visualization, centroiding)

xcms: LC-MS data pre-processing (peak detection, alignment, and correspondence)

# Data import and representation

Data from pooled human serum samples measured with a UHPLC system coupled with a Q-TOF MS instrument. The data is limited to m/z range 105-134 and retention time 0-260 seconds.

> pd is a dataframe to input sample information, such as grouping
>
> When pd is added to the object, it should be convert to class "NAnnotatedDataFrame"
>
> varMetadata can be entered for each sample information.

```{r loadData}
## Define the file names. Find them from the package directory
fls <- dir(system.file("sciex", package = "msdata"), full.names = TRUE)

# ## Define a data.frame with additional information on the files.
# pd: a 2*4 dataframe with sample information
pd <- data.frame(file = basename(fls),
                 injection_idx = c(1, 19),
                 sample = c("POOL_1", "POOL_2"),
                 group = "POOL")

data <- readMSData(fls, # filenames to load
                   # Stores data; stores sample info into @phenoData@data
                   # More info about each variable: file, injection. sample etc. can be entered in varMetadata
                   pdata = new("NAnnotatedDataFrame", pd), 
                   mode = "onDisk" # Only load metadata into memory and leave the m/z ratio vs intensity peaks on disk
                   ) 

```

# Set up parallel processing

> Set multicoreParam/SnowParam to empty unlimit the number of cores used, or enter a number for a specific amount

```{r parallel}
if (.Platform$OS.type == "unix") {
    register(bpstart(MulticoreParam(6)))
} else {
    register(bpstart(SnowParam()))
}
```

# Data object manipulation

## Phenotype data

> p stands for phenotype, which are all sorts of sample information. You can add more by editing the pd dataframe used when creating the object.
>
> pData can be used to get or set sample information (the pd matrix) directly.
>
> pData: phenoData \> data
>
> Specific sample information can be extracted directly with the \$ operator

```{r pData}
pData(data)

data$injection_idx # from data directly, not pData
```

## Feature data

> f stands for features, which are peaks in metabolomics data
>
> fData gets general information for each spectrum
>
> fData = featureData \> data

```{r fData}
kable(fData(data)[1:4,] %>% select(spectrum))
```

# Data access and visualization

## Spectra overview

> Information is organized by spectrum in xcms, which are snapshots of intensity vs. m/z diagrams in time.
>
> Functions such as `msLevel`, `centroided`, `polarity`, or `rtime` can be used to extract the respective information from all spectra.

```{r msLevel}
# MS 1 or 2
head(msLevel(data))
```

```{r centroided}
# Centroided or not, currently all not centroided
head(centroided(data))
```

```{r polarity}
# Polarity information; all positive for this example
table(polarity(data))
```

```{r rtime}
# Retention time
# split by File origin (we had 2 samples)
rts <- split(rtime(data),fromFile(data))
```

## Data filtering

> The real spectra files (not the summary, the data) need to be loaded with the `spectra` function for processing, but they are huge.
>
> So we filter data to save computational resources and memory.
>
> Filter options: `filterFile`, `filterRt`, `filterMz` or `filterMsLevel`
>
> The filter functions can be chained into dplyr chains.
>
> This will retrieve a list containing the spectra fitting the criteria

```{r filter}

sps <- data %>%
  filterRt(rt = c(180,181)) %>%
  spectra
```

> `fromFile` tells which file does the spectrum come from
>
> sps contains 3 spectra from file 1 and 3 spectra from file 2

```{r filterOverview}

length(sps)

vapply(sps, fromFile, integer(1)) # vapply requires specifying datatype; here integer(1)
```

## Plot spectrum

```{r plotSpectrum}

plot(sps[[6]]) # 3rd spectra from the 2nd file
```

## Get total ion chromatogram (TIC)

> MS data is generally organized by spectra. i.e. an intensity vs. m/z ratio photo for each timepoint
>
> The `chromatogram` function extracts the TIC, which is a intensity vs. retention time plot for each file.
>
> This will create a chromatogram object that can be filtered with `rtime` and `intensity`

```{r chromatogram-tic, fig.cap = "Total ion chromatogram.", fig.width = 10, fig.height = 5}
#' Get chromatographic data (TIC) for an m/z slice
chr <- chromatogram(data)
chr

#' Plot the tic
plot(chr)
```

The chromatogram object is a data slice x sample matrix. Here we have a 1 data slice by 2 samples matrix.

```{r}
dim(chr)
```

Chromatograph for sample 1 can be ploted as such, similarly for sample 2.

```{r}
plot(chr[1,1])
```

```{r chromatogram-tic-intensity}
ints <- intensity(chr[1, 1])
head(ints) 
```

pData can still be used to get phenotype data (sample data)

```{r chromatogram-pdata}
#' Access the full phenotype data
pData(chr)
```

There are different ways to aggregate intensity values across the mz range for the same retention time. The default is "sum", here we use "max", which gets only the max intensity m/z for each time point.

It can also be used to generate extracted ion chromatogram (EIC), which is chromatogram for a specific m/z range.

The area of under the peak corresponds to the amount of the ions present.

```{r serine-xic, fig.cap = "Extracted ion chromatogram for the Serine [M+H]+ ion in both files."}
#' Extract and plot the XIC for Serine
data %>%
    filterRt(rt = c(175, 189)) %>%
    filterMz(mz = c(106.02, 106.07)) %>%
    chromatogram(aggregationFun = "max") %>%
    plot() 
```

# Centroiding of profile MS data

Profile mode of MS data has intensities for all m/z values. To reduce file size and help peak detection, centroiding should be performed.

The `MSnbase` package provides all tools to perform centroiding (and data smoothing) in R: `pickPeaks` and `smooth`.

The top graph is the intensity vs. retention time chromatograph for serine; the bottom is m/z ratio vs. retention time, and a 3rd dimension of intensity is represented by colors. From the graph you can see that for each time point, peaks for a range of m/z ratios are detected, but the serine peak has the highest intensity between 106.02 and 106.07.

Each column of data on the bottom graph represents the data recorded in the spectrum for that timepoint.

```{r serine-profile-mode-data, fig.cap = "Profile data for Serine.", fig.width = 10, fig.height = 5, fig.pos = "h!"}
#' Filter the MS data to the signal from the Serine ion and plot it using
#' type = "XIC"
data %>%
    filterRt(rt = c(175, 189)) %>%
    filterMz(mz = c(106.02, 106.07)) %>%
    plot(type = "XIC") 
```

The data is smoothed with the SavitzkyGolay filter. `halfWindowSize` should be less than FWHM (full width at half maximum) of the peaks. Smoothing normally applies to the m/z dimension. Additionally, `combineSpectraMovingWindow` can be used to smooth both the retention time and m/z dimensions - essentially a square filter used in computer vision.

The result is a `MSnExp` object that is saved to the memory. In real use cases this will blow up the memory. So it is better to apply smoothing to each experiment file. Then export the result separately as mzML files using `writeMSData`.

Then the data is centroided by picking the max point in each spectrum. There are two other methods available: `refineMz = "descendPeak"` and `refineMz = "kNeighbors"`. Both of them compute the centroid with an intensity-weighted average of other m/z ratios in the same spectrum. KN uses 2 neighbors left and right to compute the average by default, while DP uses all the neighbors above 33% of the centroid's intensity by default before the next peak appears on left or right.

```{r centroiding, fig.cap = "Centroided data for Serine.", fig.width = 10, fig.height = 5, fig.pos = "h!"}
#' Smooth the signal, then do centroiding.
data_cent <- data %>%
    smooth(method = "SavitzkyGolay", halfWindowSize = 4) %>%
    combineSpectraMovingWindow() %>%
    pickPeaks(refineMz = "descendPeak")

#' Plot the centroided data for Serine
data_cent %>%
    filterRt(rt = c(175, 189)) %>%
    filterMz(mz = c(106.02, 106.07)) %>%
    plot(type = "XIC") 
```

However, the centroided data is not save on disk yet. Now we save them and read them in again for further manipulation.

```{r export-centroided-prepare, echo = FALSE, results = "hide"}
#' Silently removing exported mzML files if they do already exist.
lapply(basename(fileNames(data)), function (z) {
    if (file.exists(z))
        file.remove(z)
}) 
```

```{r export-centroided}
#' Write the centroided data to files with the same names in the current
#' directory
fls_new <- basename(fileNames(data))
writeMSData(data_cent, file = fls_new)
```

# Pre-processing of MS data

Pre-processing of data quantifies the signal from ion species and outputs the sample x peak matrix for downstream analysis. There are 3 steps:

1.  Peak detection

2.  Peak alignment (retention time correction)

3.  Correspondence (peak grouping)

```{r}
#' Read the centroided data.
fls_new <- c("20171016_POOL_POS_1_105-134.mzML", "20171016_POOL_POS_3_105-134.mzML")
data_cent <- readMSData(fls_new, pdata = new("NAnnotatedDataFrame", pd),
                        mode = "onDisk") 
```

## Peak Detection

Peak detection is performed with the `findChromPeaks` function on a chromatogram object. It has three methods:

1.  `MatchedFilterParam`: Performs peak detection with the vanilla paper's method

2.  `CentWaveParam`: detect peaks with continuous wavelet transformation. Two parameters `peakwidth` and `ppm` are particularly important: `peakwidth` defines the min and max width of a peak in seconds. This can be determined from internal standards or known compounds in the sample. `ppm` defines the max allowed m/z deviation in consecutive scans in ppm.

3.  `MassifquantParam`: a Kalman filter-based peak detection

### Peak width optimization

```{r centWave-default, fig.cap = "XIC for Serine", results = "hide"}
#' Get the XIC for serine in all files
srn_chr <- chromatogram(data_cent, 
                        rt = c(164, 200), # Retention time range
                        mz = c(106.03, 106.06), # mz range
                        aggregationFun = "max") # get the max
#' Plot the data
par(mfrow = c(1, 1), mar = c(4, 4.5, 1, 1))
plot(srn_chr)

#' Get default centWave parameters
cwp <- CentWaveParam()

#' "dry-run" peak detection on the XIC.
res <- findChromPeaks(srn_chr, param = cwp)
chromPeaks(res)
```

No peaks found, but this makes sense if we look at the parameters of cwp:

```{r centWave-default-parameters}
cwp 
```

Since the peaks for serine are about 4 seconds wide, we adapt the parameters below.

`integrate` defines whether peak limit is found on filtered data (1) or real data (2). Using filtered data is more robust, but less exact. Using real data is more accurate, but prone to noise.

There are two peaks because there are two data files.

```{r centWave-adapted, fig.cap = "XIC for Serine with detected chromatographic peak", results = "hide"}
cwp <- CentWaveParam(peakwidth = c(2, 5), integrate = 2)

srn_chr <- findChromPeaks(srn_chr, param = cwp)

#' Plot the data and higlight identified peak area
plot(srn_chr)
```

Identified peaks can be viewed with `chromPeaks` function.

`rt`, `rtmin`, `rtmax`: retention time

`into`: integrated peak area

`intb`: base-line corrected integrated peak area

`maxo`: maximum signal

`sn`: s/n ratio

```{r chromPeaks-chromatogram}
chromPeaks(srn_chr)
```

A matrix to annotate the identified peaks can be extracted as well.

```{r}
chromPeakData(srn_chr)
```

### Ppm optimization

The ppm defines how much variation in ppm is allowed for a peak to be considered valid. The higher the intensity, the smaller the variation should be. This parameter should also be tuned.

First get the centroided serine data.

Plot the XIC for srn:

```{r Serine-mz-scattering-plot}
#' Restrict the data to signal from Serine
srn <- data_cent %>%
    filterRt(rt = c(179, 186)) %>%
    filterMz(mz = c(106.04, 106.06))

#' Plot the data
plot(srn, type = "XIC") 
```

I did some fancy centroiding so the variation does not exactly match the description above, but in general the variation is less when intensity is high.

Next we find the sample (we have 2) with the highest intensity overall in the serine range. This will make a stricter requirement for ppm later on.

```{r sample-highest-signal}
#' Extract intensities and split them by file. This will return
#' a list of lists.
ints_per_file <- split(intensity(srn), fromFile(srn))

#' For each file, sum up the intensities.
ints_sum <- lapply(ints_per_file, function(x) sum(unlist(x)))
ints_sum
```

In order to compute variations, we need the difference in intensity between consecutive scans in time.

`filterFile` operates on a MSnExp object to get the one with the highest total intensity. `as` converts it to a data.frame.

`diff` is a function that computes lagged difference. i.e. next number subtracts the previous, with more options.

```{r define-ppm}
#' Extract the Serine data for one file as a data.frame
srn_df <- as(filterFile(srn, file = which.max(ints_sum)), "data.frame")

#' The difference between m/z values from consecutive scans expressed
#' in ppm
(diff(srn_df$mz) * 1e6 / mean(srn_df$mz)) %>% summary
```

So we are setting the ppm limit from 0 to 10 based on serine. Ideally this should be tested for a few more compounds to get the optimal ppm range. The result is a XCMSnExp object which inherits from the MSnExp object and also contains all pre-processing results

```{r findPeaks-centWave}
#' Perform peak detection
cwp <- CentWaveParam(peakwidth = c(2, 5), ppm = 30, integrate = 1)
data_cent <- findChromPeaks(data_cent, param = cwp) 
```

The `chromPeaks` function can be used to access detected peaks with mz and rt as filtering criteria.

For each identified peak the values of its apex is reported (`mz`, `rt`, and their ranges).

`into`: integrated signal

`maxo`: max signal

`sn`: s/n ratio

`sample`: index

```{r xcmsnexp}
#' Access the peak detection results from a specific m/z - rt area
chromPeaks(data_cent, mz = c(106, 107), rt = c(150, 190))
```

Extracted ion chromatogram will also contain detected peaks after peak detection. Here the EIC of serine:

```{r chrom-after}
eic_serine <- chromatogram(data_cent, mz = c(106.04, 106.06),
                           rt = c(179, 186))
chromPeaks(eic_serine)
```

Plot the chromatogram

```{r plot-after}
plot(eic_serine)
```

An XIC plot after peak detection also shows the rt/mz space from which the peak signal is integrated.

```{r xic-after}
srn <- data_cent %>%
    filterRt(rt = c(175, 188)) %>%
    filterMz(mz = c(106.04, 106.06))

plot(srn, type = "XIC")
```

### Peak post-processing

Two issues need to be handled: overlapping peaks and artificially split peaks. This is done with the `refineChromPeaks` function.

For the former, a `cleanPeaksParam` can be passed to the function. For the latter, a `MergeNeighboringPeaksParam` object can be manipulated.

Merging overlap peaks: if the peaks overlap for 4 seconds and the signal between them is lower than 75% of the smaller peak's max intensity.

Weird parallel computing issues sometimes happen.

```{r, results="hide"}
mpp <- MergeNeighboringPeaksParam(expandRt = 4)
data_cent_pp <- refineChromPeaks(data_cent, param = mpp)
```

Whether a peak is merged can be found via the `chromPeakData` function.

```{r}
chromPeakData(data_cent_pp)
```

```{r merged-peak, fig.width = 12, fig.height = 6, fig.cap = "Result from the peak refinement. Left: peaks before merging, right after merging."}
mzr <- c(124.084, 124.088)
rtr <- c(150, 170)
chr_1 <- chromatogram(filterFile(data_cent, 2), mz = mzr, rt = rtr)
chr_2 <- chromatogram(filterFile(data_cent_pp, 2), mz = mzr, rt = rtr)
par(mfrow = c(1, 2))
plot(chr_1)
plot(chr_2)
```

```{r}
data_cent <- data_cent_pp
```

```{r plotChromPeaks, fig.cap = "Location of the identified chromatographic peaks in the m/z - rt space.", fig.height = 7, fig.width = 12}
par(mfrow = c(1, 2))
plotChromPeaks(data_cent, 1)
plotChromPeaks(data_cent, 2) 
```

## Peak Alignment

```{r alignment-bpc-raw, fig.cap = "BPC of all files.", fig.width = 8, fig.height = 4}
#' Extract base peak chromatograms
bpc_raw <- chromatogram(data_cent, aggregationFun = "max")
plot(bpc_raw, peakType = "none")
```

```{r alignment-correspondence}
#' Define the settings for the initial peak grouping - details for
#' choices in the next section.
pdp <- PeakDensityParam(sampleGroups = data_cent$group, bw = 1.8,
                        minFraction = 1, binSize = 0.02)
data_cent <- groupChromPeaks(data_cent, pdp)

#' Define settings for the alignment
pgp <- PeakGroupsParam(minFraction = 1, span = 0.6)
data_cent <- adjustRtime(data_cent, param = pgp) 
```

```{r alignment-result, fig.width = 8, fig.height = 4, fig.cap = "Alignment results. Shown is the difference between raw and adjusted retention times and the hook peaks that were used for the alignment (shown as points)."}
#' Plot the difference between raw and adjusted retention times
plotAdjustedRtime(data_cent) 
```

```{r bpc-raw-adjusted, fig.cap = "BPC before (top) and after (bottom) alignment.", fig.width = 10, fig.height = 8}
par(mfrow = c(2, 1))
#' Plot the raw base peak chromatogram
plot(bpc_raw, peakType = "none")
#' Plot the BPC after alignment
plot(chromatogram(data_cent, aggregationFun = "max", include = "none"))
```

```{r serine-xic-adjusted, fig.cap = "XIC for Serine before (left) and after (right) alignment", fig.width = 10, fig.height = 4}
#' Use adjustedRtime parameter to access raw/adjusted retention times
par(mfrow = c(1, 2), mar = c(4, 4.5, 1, 0.5))
plot(chromatogram(data_cent, mz = c(106.04, 106.06),
                  rt = c(179, 186), adjustedRtime = FALSE))
plot(chromatogram(data_cent, mz = c(106.04, 106.06),
                  rt = c(179, 186))) 
```

## Correspondence

```{r correspondence-example, fig.cap = "BPC for a m/z slice and defined features within this slice based on default settings.", fig.width = 10, fig.height = 7}
#' Get default parameters for the grouping
pdp <- PeakDensityParam(sampleGroups = data_cent$group)

#' Extract a BPC for the m/z slice containing serine
bpc_serine <- chromatogram(data_cent, mz = c(106.04, 106.06),
                           aggregationFun = "max")

#' Dry-run correspondence and show the results.
plotChromPeakDensity(bpc_serine, param = pdp)
```

```{r correspondence-bw, fig.cap = "Correspondence analysis with default settings on an m/z slice containing signal from multiple ions.", fig.width = 10, fig.height = 7}
#' Plot the chromatogram for an m/z slice containing Betaine and Valine
mzr <- 118.08625 + c(-0.01, 0.01)
chr <- chromatogram(data_cent, mz = mzr, aggregationFun = "max")

#' Correspondence in that slice using default settings
pdp <- PeakDensityParam(sampleGroups = data_cent$group)
plotChromPeakDensity(chr, param = pdp)
```

```{r correspondence-bw-fix, fig.cap = "Correspondence analysis with reduced bw setting on a m/z slice containing signal from multiple ions.", fig.width = 10, fig.height = 7}
#' Reducing the bandwidth
pdp <- PeakDensityParam(sampleGroups = data_cent$group, bw = 1.8)
plotChromPeakDensity(chr, param = pdp)
```

```{r correspondence-analysis}
pdp <- PeakDensityParam(sampleGroups = data_cent$group, bw = 1.8,
                        minFraction = 0.4, binSize = 0.02)

#' Perform the correspondence analysis
data_cent <- groupChromPeaks(data_cent, param = pdp) 
```

```{r correspondence-evaluate, fig.cap = "Result of correspondence on a slice containing the isomers Leucine and Isoleucine.", fig.width = 10, fig.heigt = 7}
#' Plot the results for an m/z slice containing Leucine and Isoleucine
mzr <- 132.10191 + c(-0.01, 0.01)
chr <- chromatogram(data_cent, mz = mzr, aggregationFun = "max")
plotChromPeakDensity(chr, simulate = FALSE)
```

```{r correspondence-featureDefinitions}
#' Definition of the features
featureDefinitions(data_cent) 
```

```{r correspondence-featureSummary}
#' Per-feature summary.
head(featureSummary(data_cent)) 
```

```{r correspondence-featureValue}
#' feature intensity matrix
fmat <- featureValues(data_cent, value = "into", method = "maxint")
head(fmat)
```

```{r fillChromPeaks}
#' Number of missing values
sum(is.na(fmat))

data_cent <- fillChromPeaks(data_cent, param = ChromPeakAreaParam())

#' How many missing values after
sum(is.na(featureValues(data_cent)))

fmat_fld <- featureValues(data_cent, value = "into", method = "maxint")
head(fmat_fld) 
```

```{r}
library(SummarizedExperiment)
res <- quantify(data_cent, filled = FALSE, method = "sum", value = "into")
```

```{r}
colData(res)
```

```{r}
colData(res)
```

```{r}
head(assay(res, "raw"))
```

```{r}
assays(res)$raw_filled <- featureValues(data_cent, method = "sum",
                                        value = "into", filled = TRUE)
```

```{r}
assayNames(res)
```

```{r correspondence-result-object}
#' Overview of the performed processings
processHistory(data_cent)
```

```{r correspondence-history}
#' Access the parameter class for a processing step
processParam(processHistory(data_cent)[[1]])
```

```{r}
metadata(res)
```
